---
title: "Comparing Schoendfeld vs. Freedman formulas for sample size"
author: "Jordi Cortés"
date: "18/2/2020"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(survival))
suppressMessages(library(knitr))
suppressMessages(library(snowfall))

# Sample size formulas
schoendfeld.formula <- function(alpha,power,HR) E <- 4*(qnorm(1-alpha/2) +  qnorm(power))^2 / (log(HR))^2
freedman.formula    <- function(alpha,power,HR) E <- (HR+1)^2 * (qnorm(1-alpha/2) + qnorm(power))^2/(HR-1)^2

##-- Log-rank function for vectorize
f.log.rank <- function(t,tau,Treatment){
  STATUS <- t<tau
  t[t>=tau] <- tau 
  LR <- survdiff(Surv(t,STATUS)~Treatment)
  return(1 - pchisq(LR$chisq,1))
}


# Panalize scientific notation
options(scipen=5)

# sfInit for parallel computing
sfInit(parallel=TRUE, cpus=7)
suppressMessages(sfLibrary(survival))
```

## Simulations

This document has the aim to shown the performance of **Freedman** and **Schoendfeld** formulas by a simple simulation.

- Schoendfeld formula:

$E=\frac{4(Z_{\alpha/2}+Z_{\beta})^2}{\left(log(HR)\right)^2}$

- Freedman formula:

$E=\frac{(HR+1)^2(Z_{\alpha}+Z_{\beta})^2}{\left(HR-1\right)^2}$

```{r simulations,message=FALSE,warning=FALSE}
set.seed(12345)                  # Seed to replicate the results
nsim <- 10000                    # number of simulations
HR <- seq(0.1,0.95,0.05)         # HR tested
power_S <- power_F <- c()        # Store powers of Schoendfeld (S) and Freedman (F) formulas.

for(hr in HR){
  
  ##-- Schoendfeld-----------------------------
  E <- schoendfeld.formula(alpha=0.05,power=0.8,HR=hr)
  N <- 2*ceiling(E/(pexp(1,1)+pexp(1,hr)))
  Treatment <- c(rep(0,N/2),rep(1,N/2))
  
  # Generate times
  time0_C <- matrix(rexp(nsim*N/2,1),nrow=nsim)
  time0_T <- matrix(rexp(nsim*N/2,hr),nrow=nsim)
  time <- cbind(time0_C,time0_T)
  p <- sfApply(time,1,f.log.rank,tau=1,Treatment=Treatment)
  power_S[which(HR==hr)] <- sum(p<0.05)/nsim

  
  
  ##-- Freedman -------------------------------
  E <- freedman.formula(alpha=0.05,power=0.8,HR=hr)
  N <- 2*ceiling(E/(pexp(1,1)+pexp(1,hr)))
  Treatment <- c(rep(0,N/2),rep(1,N/2))
  
  # Generate times
  time0_C <- matrix(rexp(nsim*N/2,1),nrow=nsim)
  time0_T <- matrix(rexp(nsim*N/2,hr),nrow=nsim)
  time <- cbind(time0_C,time0_T)
  p <- sfApply(time,1,f.log.rank,tau=1,Treatment=Treatment)
  power_F[which(HR==hr)] <- sum(p<0.05)/nsim
}
sfStop()
```

The simulation parameters are the following:

- **Number of simulations**:`r nsim`
- **Follow-up &Tau;**: 1
- **HRs tested**:`r paste0(HR,collapse=', ')`
- **Test**: Log-rank
- **Exponential distributions**:

$T_C \sim Exp(\lambda_C=1)$

$T_T \sim Exp(\lambda_T=HR)$

## Results

These plots summarize the results.

```{r fig.width=10}
par(mfrow=c(1,3))
plot(HR,power_S,type='b',ylim=c(0.70,0.90));abline(h=0.8,lty=2)
plot(HR,power_F,type='b',ylim=c(0.70,0.90));abline(h=0.8,lty=2)
plot(HR,power_F/power_S,type='b');abline(h=1,lty=2)
```

The next table contains the power achieved for each formula:

```{r}
kable(data.frame(HR,power_S,power_F),digits=3,col.names = c('Hazard Ratio','Power Schoendfeld','Power Freedman'))
```


**Conclusion**: The performance of Schoendfeld is quite poor with HRs far away of 1. Moreover, the number of events can differ between two methods up to 30% [1].

## References

1. Ulrich R. Abel, Katrin Jensen, Irini Karapanagiotou-Schenkel and Meinhard Kieser. Some issues of sample size calculation for time-to-event endpoints using the Freedman and Schoenfeld formulas. Journal of Biopharmaceutical Statistics. (DOI: 10.1080/10543406.2014.1000546)[http://dx.doi.org/10.1080/10543406.2014.1000546]
